The Galactic Midplane Is Not a Plane: Implications for Dynamical Analysis with {\em Gaia} Data and Beyond

### 2019-04-30 paper/ms.tex
Fix eq ref in Fig 4 caption
Update author information
Remove all mentions of clusters

\author{Angus Beane}
\affil{\cca}
\affil{\penn}

\author{Robyn Sanderson}
\affil{\cca}
\affil{\penn}

@@

\author{Douglas Filho}
\affil{\columbia}

\author{Mordecai-Mark Mac Low}
\affil{\cca}
\affil{\amnh}

@@

\author{David W. Hogg}
\affil{\nyuphys}
\affil{\nyucds}
\affil{\cca}
\affil{\mpia}

@@

\keywords{Galaxy:~disk -- solar~neighborhood --
open~clusters~and~associations:~general -- Galaxy:~structure --
Galaxy:~evolution}

@@

In Section~\ref{sec:discussion} we discuss the implications of midplane
variations, and the resulting systematic uncertainty in the vertical action,
for action-space analyses beyond the dynamical solar neighborhood. In light of
our findings, we considered the orbits of the nine open clusters within
$250\,\pc$ as reported by the {\em Gaia} collaboration
\citep{2018AA...616A..10G}. In Table~\ref{tab:real_clusters}, we report the
distance to each cluster, the three actions, and the maximum vertical extent
of their orbits ($z_{\text{max}}$). In Section~\ref{sssec:reconstruction} we
estimate and discuss the size of the dynamical solar neighborhood for each
cluster based on its mass, to show that action-space tagging is likely still a
viable method if contamination from foreground or background stars is
unimportant. We summarize and conclude in Section~\ref{sec:conclusion}.

\begin{deluxetable*}{cCCCCCCCC}
\tablecaption{Relevant dynamical parameters for the $9$ open clusters within
$250\,\pc$ of the Sun, using the relevant initial positions and velocities
from \citet{2018AA...616A..10G}. Actions are computed using the method
outlined in Section~\ref{ssec:action_comp}, with typical values for the
Galactocentric coordinate system parameters. These are only shown to
illustrate the general types of orbits open clusters are
on.\label{tab:real_clusters}}
\tablehead{\colhead{cluster} & \colhead{$N_{\text{memb}}$} &
\colhead{distance} & \colhead{$J_R$} &
\colhead{$J_{\phi}$} & \colhead{$J_z$} & \colhead{$z_{\text{max}}$} &
\colhead{mass} \\
\colhead{ } & \colhead{ } &
\colhead{$\mathrm{pc}$} & \colhead{$\mathrm{kpc\,km\,s^{-1}}$} &
\colhead{$\mathrm{kpc\,km\,s^{-1}}$} & \colhead{$\mathrm{kpc\,km\,s^{-1}}$} &
\colhead{$\mathrm{pc}$} & \colhead{$M_{\odot}$}}
\startdata
alphaPer & 740 & 170 & 20 & -1700 & 0.005 & 11 & 352\tablenotemark{b} \\
Blanco1 & 489 & 240 & 2 & -1800 & 2 & 210 & 410\tablenotemark{c} \\
ComaBer & 153 & 90 & 2 & -1900 & 0.7 & 140 & 112\tablenotemark{d} \\
Hyades & 515 & 50 & 20 & -1800 & 0.3 & 80 & 400\tablenotemark{e} \\
IC2391 & 325 & 150 & 6 & -1800 & 0.02 & 20 & \nodata \\
IC2602 & 492 & 150 & 10 & -1700 & 0.3 & 89 & \nodata \\
NGC2451A\tablenotemark{a} & 400 & 190 & 6 & -1800 & 0.2 & 73 & \nodata \\
Pleiades & 1326 & 140 & 20 & -1700 & 0.4 & 96 & 800\tablenotemark{f} \\
Praesepe & 938 & 190 & 20 & -1800 & 0.6 & 120 & 550\tablenotemark{d}
\enddata

\tablenotetext{a}{This cluster is labeled NGC2451, which we interpret as
NGC2451A since NGC2451B lies further than $250\,\pc$ away.}
\tablenotetext{b}{\citet{2016MNRAS.457.1028S}}
\tablenotetext{c}{\citet{2007AA...471..499M}}
\tablenotetext{d}{\citet{2007AJ....134.2340K}}
\tablenotetext{e}{\citet{1998AA...331...81P}}
\tablenotetext{f}{\citet{2001AJ....121.2053A}}

\tablerefs{\citet{2018AA...616A..10G}}
\end{deluxetable*}

@@

Figure~\ref{fig:one_orbit_wrong_ref} shows that the actions computed in the
offset coordinate systems depend on the time (i.e. orbital phase) at which the
star's phase-space position is observed. To the immortal observer this would
appear as time-dependence in the actions, which appear to oscillate around
their value using the correct coordinate system even though in this example
the observer is using the correctly constructed best-fit static, axisymmetric
potential. The relative size of the phase variation in each action depends on
the direction of the systematic offset as well as the true values of the
actions (i.e. the type of orbit). In reality we will have one measurement of
the phase-space position to work with, in which case the determination of the
orbital phase in $z$ or $R$ is degenerate with the degree of systematic offset
in that coordinate (see Figure~\ref{fig:cartoon}). Additionally, in searching
for migrated members of known open clusters, we would compute actions for
stars with a wide range of orbital phases and locations in the Galaxy. In the
following we therefore quote percentile ranges for the possible values
computed for each action as a proxy for the effect of these systematic errors
in the coordinate system.

@@

\begin{figure*}
\begin{center}
\includegraphics[width=\textwidth]{fig/schmactions_many_orbits.pdf}
\end{center}
\caption{We report one half the 95\uth{} minus 5\uth{} percentile of the error
in the measured action ($\Delta J_i$) from coordinate system errors for the
thin, thick, and halo orbits (Table~\ref{tab:orbits}). See discussion in the
text for the justification in using this to measure the magnitude of the
induced error. The {\em left}, {\em center}, and {\em right} panels show the
result for $J_R$, $J_{\phi}$, and $J_z$, respectively. The {\em upper} panels
consider an offset in $z$ and the {\em lower} panels consider an offset in $x$
(equivalently, an offset in the Solar radius). In some panels we also plot as
dashed lines the epicyclic prediction of the induced action error
(Equations~\eqref{eq:actions_epi_error}~and~\eqref{eq:Jr_epi_dx}). In the
epicyclic approximation, a $z$ offset only induces an error in $J_z$ --- for
all three orbits the epicyclic approximation is a good description of the
$J_z$ error. An $x$ offset induces an error in $J_R$ and $J_{\phi}$. The error
in $J_R$ is somewhat well-described for the thin-disk and thick-disk orbits,
and a poor description for the halo orbit. For $J_{\phi}$, the epicyclic
approximation is not a good description for any orbit.}
\label{fig:many_orbit_wrong_ref}
\end{figure*}

@@

This finding has many implications for the study of the dynamics of stars in
what is normally considered the regime of the epicyclic approximation:
quasi-harmonic oscillations around a circular ``guiding center orbit.'' Here
we consider a few of those implications, starting with an estimate of the
region around the Sun where we expect an extension of the local axisymmetric
coordinate system, and therefore the epicyclic approximation, to still be
valid. This region is also where it should be feasible to perform ``dynamical
tagging'' in the thin disk; that is, using the computed orbits or integrals of
motion of field stars to match them with known open clusters.

\subsection{Disrupted Cluster Reconstruction}
\label{sssec:reconstruction}

We now consider the implications for attempts to reconstruct disrupting open
clusters by selecting ejected members based on their actions.

Suppose an open cluster of mass $m_c$ on an orbit with actions $\bm{J}_c$ is
disrupted. We would like to determine which stars with complete 6D phase-space
positions, for example in the {\em Gaia} dataset, are likely to be ejected
members from that cluster.\footnote{This is likely to be more successful with
DR3/4 than DR2, since upcoming data releases will contain a far greater number
of stars with radial velocities --- though targeted, follow-up radial velocity
surveys may allow this program to be performed sooner.} One could generate an
initial catalogue of candidate members by integrating the cluster's orbit and
selecting stars close to that orbit, accounting for the fact that the stream
of ejected stars will not exactly follow the orbit of the cluster
\citep[e.g.][]{2011MNRAS.413.1852E}.

One could then make a further selection by requiring that the action of each
field star $i$ be close to the cluster's actions to within some bound, i.e.
$\bm{J}_i \in V_e$, where $V_e$ is a volume centered on $\bm{J}_c$ and has
size comparable to the intrinsic action-space dispersion of the disrupting
cluster. This dispersion can be estimated as
\citep[Section~8.3.3][]{2008gady.book.....B},
\beq \label{eq:action_disp}
\frac{\Delta J_i}{J_i} \sim \left(\frac{m_c}{M}\right)^{1/3}\text{,}
\eeq
where $m_c$ is the mass of the cluster and $M$ is the mass enclosed by the
cluster's orbit around the Galaxy, in this case assumed to be the mass inside
the Solar circle ($\sim 10^{11}\Msun$). Stars with actions within $V_e$ would
then be considered cluster members.

Two issues are immediately apparent with this approach. The first is that it
neglects the overall smooth distribution \emph{in action space} which will
contribute some stars to the selection by chance. For dissolving open clusters
in the thin disk, this background is likely to be significant, but we will
assume for the sake of argument that these foreground stars can be modeled
within the ``dynamical solar neighborhood'' where it is safe to extend the
local coordinate system. Since the effect of extending the coordinate frame
beyond this region is generally to inflate the actions computed for thin-disk
stars, using too large a volume is likely to under-estimate the density of the
smooth thin-disk-like component in action space, while scattering more
thin-disk stars onto thick-disk-like orbits and over-estimating the density of
the smooth thick-disk-like component.

The second issue, which is related to the point of this paper, is that based
on the results in our previous sections computing $\bm{J}_i$ for a field star
and directly comparing it to  $\bm{J}_c$ will have significant systematic
error if the cluster and field star are far enough apart to have significantly
different local midplanes (see
Figures~\ref{fig:midplane}~and~\ref{fig:midplane_chervin}). Since our
criterion for cluster membership (distance in action space) scales with the
cluster mass, the region where the search will succeed varies with cluster
mass as well. To illustrate this effect, we assume a cluster mass $m_c$ and
estimate the expected action-space dispersion from
Equation~\eqref{eq:action_disp}. We then consult
Figure~\ref{fig:many_orbit_wrong_ref} to determine the midplane offset
necessary in order for the induced error in the vertical action to be the same
as the expected action-space dispersion. We perform this analysis for the
thin-disk-like and thick-disk-like orbits considered in
Section~\ref{ssec:quant}, assuming that the halo-like orbit is unlikely for
members of known open clusters (see Table~\ref{tab:real_clusters}). The result
for the allowable midplane offset for these two orbits in shown in
Figure~\ref{fig:cluster_offset} as a function of cluster mass.

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{fig/cluster_offset.pdf}
\end{center}
\caption{The $z$ offset necessary in order for the action error induced by our
midplane effect to be comparable to the intrinsic action dispersion from a
cluster of mass $m_c$. We compute the intrinsic dispersion from
Equation~\eqref{eq:action_disp} and convert the dispersion to a $z$ offset using
the result from Figure~\ref{fig:many_orbit_wrong_ref}. We consider the result
for the thin and thick orbit (Table~\ref{tab:orbits}). Nearby open clusters
(Table~\ref{tab:real_clusters}) have orbits closest to the thin orbit.}
\label{fig:cluster_offset}
\end{figure}

>>>>

\author[0000-0002-8658-1453]{Angus Beane}
\affil{\cca}
\affil{\penn}

\author[0000-0003-3939-3297]{Robyn Sanderson}
\affil{\cca}
\affil{\penn}

@@

\author[0000-0001-8275-9181]{Douglas Grion Filho}
\affil{\columbia}

\author[0000-0003-0064-4060]{Mordecai-Mark Mac Low}
\affil{\cca}
\affil{\amnh}

@@

\author[0000-0003-2866-9403]{David W. Hogg}
\affil{\nyuphys}
\affil{\nyucds}
\affil{\cca}
\affil{\mpia}

@@

\keywords{Galaxy:~disk -- solar~neighborhood -- Galaxy:~structure --
Galaxy:~evolution}

@@

In Section~\ref{sec:discussion} we discuss the implications of midplane
variations, and the resulting systematic uncertainty in the vertical action,
for action-space analyses. In particular, we consider the impact on modelling
the distribution function of the Milky Way in Section~\ref{ssec:df}. We
summarize and conclude in Section~\ref{sec:conclusion}.

@@

Figure~\ref{fig:one_orbit_wrong_ref} shows that the actions computed in the
offset coordinate systems depend on the time (i.e. orbital phase) at which the
star's phase-space position is observed. To the immortal observer this would
appear as time-dependence in the actions, which appear to oscillate around
their value using the correct coordinate system even though in this example
the observer is using the correctly constructed best-fit static, axisymmetric
potential. The relative size of the phase variation in each action depends on
the direction of the systematic offset as well as the true values of the
actions (i.e. the type of orbit). In reality we will have one measurement of
the phase-space position to work with, in which case the determination of the
orbital phase in $z$ or $R$ is degenerate with the degree of systematic offset
in that coordinate (see Figure~\ref{fig:cartoon}).  In the following we
therefore quote percentile ranges for the possible values computed for each
action as a proxy for the effect of these systematic errors in the coordinate
system.

@@

\begin{figure*}
\begin{center}
\includegraphics[width=\textwidth]{fig/schmactions_many_orbits.pdf}
\end{center}
\caption{We report one half the 95\uth{} minus 5\uth{} percentile of the error
in the measured action ($\Delta J_i$) from coordinate system errors for the
thin, thick, and halo orbits (Table~\ref{tab:orbits}). See discussion in the
text for the justification in using this to measure the magnitude of the
induced error. The {\em left}, {\em center}, and {\em right} panels show the
result for $J_R$, $J_{\phi}$, and $J_z$, respectively. The {\em upper} panels
consider an offset in $z$ and the {\em lower} panels consider an offset in $x$
(equivalently, an offset in the Solar radius). In some panels we also plot as
dashed lines the epicyclic prediction of the induced action error
(Equation~\eqref{eq:Ji_err_mosttime}). In the epicyclic approximation, a $z$
offset only induces an error in $J_z$ --- for all three orbits the epicyclic
approximation is a good description of the $J_z$ error. An $x$ offset induces
an error in $J_R$ and $J_{\phi}$. The error in $J_R$ is somewhat
well-described for the thin-disk and thick-disk orbits, and a poor description
for the halo orbit. For $J_{\phi}$, the epicyclic approximation is not a good
description for any orbit.}
\label{fig:many_orbit_wrong_ref}
\end{figure*}

@@

This finding has many implications for the study of the dynamics of stars in
what is normally considered the regime of the epicyclic approximation:
quasi-harmonic oscillations around a circular ``guiding center orbit.'' Here
we consider those implications for the prospect of distribution function
fitting.

\subsection{Distribution Function Fitting}
\label{ssec:df}
